---
title: "Assignment 10: Data Scraping"
author: "Cecilia Liang"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
#Install familiar packages
library(tidyverse);library(lubridate);library(viridis);library(here)
here()

#install.packages("rvest")
library(rvest)

#install.packages("dataRetrieval")
library(dataRetrieval)

#install.packages("tidycensus")
library(tidycensus)

# Set theme
mytheme <- theme_gray() +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
webpage_URL <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024')
webpage_URL
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

water_system_name <- webpage_URL %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text()
water_system_name

PWSID <- webpage_URL %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text()
PWSID

Ownership <- webpage_URL %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text()
Ownership

MGD <- webpage_URL %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text()
MGD

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
df_durham <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "June", "Oct",
            "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"), 
                        "Year" = rep(2024, 12),
            "Maximum_Day_Use" = as.numeric(MGD))

#include the facility name and type as well as the date 
df_durham <- df_durham %>%
  mutate(Date = my(paste(Month, Year))) %>%
  arrange(Date)

#5 
ggplot(df_durham, aes(x = Date, y = Maximum_Day_Use)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Maximum Daily Withdrawals in 2024",
    subtitle = "Durham Municipal LWSP",
    x = "Month",
    y = "Maximum Daily Withdrawal ") +
  theme_minimal()

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape_it <- function(PWSID, year){
  webpage_URL <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
                              PWSID, '&year=', year))
  water_system_name <- 'div+ table tr:nth-child(1) td:nth-child(2)'
  PWSID <- 'td tr:nth-child(1) td:nth-child(5)'
  Ownership <- 'div+ table tr:nth-child(2) td:nth-child(4)'
  MGD <- 'th~ td+ td , th~ td+ td'
  water_system_name <- webpage_URL %>% html_nodes(water_system_name) %>% html_text()
  PWSID <- webpage_URL %>% html_nodes(PWSID) %>% html_text()
  Ownership <- webpage_URL %>% html_nodes(Ownership) %>% html_text()
  MGD <- webpage_URL %>% html_nodes(MGD) %>% html_text()
  df_durham <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "June", "Oct",
            "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"), 
                        "Year" = rep(year, 12),
            "Maximum_Day_Use" = as.numeric(MGD),
            "Water_System" = rep(water_system_name, 12),
            "PWSID" = rep(PWSID, 12),
            "Ownership" = rep(Ownership, 12)) %>%
    mutate(Date = my(paste(Month, Year))) %>%
    arrange(Date)
  return(df_durham)
}


```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df_durham_2020 <- scrape_it("03-32-010", 2020)

ggplot(df_durham_2020, aes(x = Date, y = Maximum_Day_Use)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Maximum Daily Withdrawals in 2020",
    subtitle = "Durham Municipal LWSP",
    x = "Month",
    y = "Maximum Daily Withdrawal") +
  theme_minimal()

glimpse(df_durham_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
df_cary_2020 <- scrape_it("03-92-020", 2020)

ggplot(df_cary_2020, aes(x = Date, y = Maximum_Day_Use)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Maximum Daily Withdrawals in 2020",
    subtitle = "Cary Municipal LWSP",
    x = "Month",
    y = "Maximum Daily Withdrawal") +
  theme_minimal()

glimpse(df_cary_2020)

df_cary_durham_2020 <- bind_rows(df_durham_2020, df_cary_2020)


ggplot(df_cary_durham_2020, aes(x = Date, y = Maximum_Day_Use, color = Water_System)) +
  geom_line() +
  labs(
    title = "Maximum Daily Withdrawals: Durham vs. Cary",
    x = "Month",
    y = "Maximum Daily Withdrawal",
    color = "Water System") +
  theme_minimal()

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
PWSID_cary <- "03-92-020"
years <- 2018:2023

df_cary_all <- map2(PWSID_cary, years, scrape_it) |> bind_rows()

ggplot(df_cary_all, aes(x = Date, y = Maximum_Day_Use)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(
    title = "Cary Maximum Daily Withdrawals (2018–2023)",
    x = "Month",
    y = "Maximum Daily Withdrawal") +
  theme_minimal()


```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer: By looking at the plot, Cary doesn't have a strong trend in water usage, while there is a small rise in 2021, a slightly upward trend was shown by the smoothed line. 
>
>
