---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Cecilia Liang"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
```{r, message = FALSE}
library(tidyverse)
library(agricolae)   
library(here)
library(ggplot2)
getwd()




#1
LTER_lake_chem <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"))


#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"),
        panel.background = element_rect(fill ="white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        plot.title = element_text(
          hjust = 0.5,         
          face = "bold"),
        legend.position = "middle")


```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature in July does not change with depth across all lakes.
Ha: Mean lake temperature in July does change with depth across all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

lake_july <- LTER_lake_chem %>%
  mutate(sampledate = as.Date(sampledate, format = "%m/%d/%Y")) %>%
  filter(format(sampledate, "%m") == "07") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

#5
tempBydepth <- 
  ggplot(lake_july,
         aes(x=depth, y=temperature_C))+
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  geom_point(alpha = 0.5, color = "blue", size = 1.5)+
  ylim(0, 35)+
  mytheme+
  labs(
    title = "Relationship Between Lake Depth and Temperature in July",
    x = "Depth (m)",
    y = "Temperature (°C)")+
  theme(
    plot.title = element_text(size=11))

print(tempBydepth)


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatterplot shows a negative relationship between lake depth and temperature in July. As depth increases, the mean lake temperature generally decreases, indicating that deeper water is cooler. However, the distribution of points shows that the relationship may not be perfectly linear: At shallow depths (around 0–3 m), temperatures vary widely, indicating strong surface variability among lakes or sampling times. At greater depths (>10 m), the temperature levels off near a lower limit, indicating possible nonlinear flattening.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
tempBydepth_regression <- lm(
  data = lake_july,
  temperature_C ~ depth)
summary(tempBydepth_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression analysis revealed a strong and statistically significant relationship between lake depth and mean temperature in July (F(1, 9726) = 2.75e+04, p < 2.2e-16). Depth explained approximately 73.9% of the variability in lake temperature (Adjusted R-squared = 0.7387), indicating that most of the variation in temperature across all lakes is associated with changes in depth. The fitted model, Temperature = 21.96 – 1.95 × Depth, suggests that for every 1 meter increase in depth, mean lake temperature decreases by about 1.95 °C. The intercept (21.96 °C) represents the estimated surface temperature, while the residual standard error of 3.84 °C indicates moderate unexplained variation, likely due to other environmental or lake-specific factors not included in the model.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
tempBydepth_AIC <- lm(data=lake_july,
                    temperature_C ~ depth + year4 + daynum)

step(tempBydepth_AIC)


#10
temp_multiregression <- lm(data=lake_july,
                    temperature_C ~ depth + year4 + daynum)

summary(temp_multiregression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC model selection procedure retained depth, year4, and daynum as the final set of explanatory variables for predicting lake temperature. This means that temperature is best explained not only by depth, but also by the year and the day of sampling within July. The final multiple regression model explains approximately 74.1% of the variance in lake temperature (Adjusted R-squared = 0.7412). Compared to the simpler model using only depth (Adjusted R-squared = 0.7387), this model represents a slight improvement in explanatory power. The inclusion of year4 and daynum adds small but statistically significant contributions, indicating that temperature varies slightly across years and throughout the month of July, in addition to changes with depth.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

temp_July_anova <- aov(temperature_C ~ lakename, data = lake_july)
summary(temp_July_anova)

# Format ANOVA as lm
temp_July_lm <- lm(temperature_C ~ lakename, data = lake_july)
anova(temp_July_lm)
summary(temp_July_lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: The analysis indicates that there is a statistically significant difference in mean temperature among the lakes (p value < 2.2e-16). This means that at least one lake’s mean July temperature differs significantly from the others.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
lake_tempBydepth <- ggplot(lake_july, 
                           aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(aes(group = lakename), method = "lm", se = FALSE) +  # add group mapping
  coord_cartesian(ylim = c(1, 35)) +  # better than ylim() to avoid dropping data
  labs(
    title = "Relationship Between Lake Depth and Temperature in July",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name"
  ) +
  mytheme +
  theme(
    plot.title = element_text(size = 11),
    legend.position = "right"
  )

print(lake_tempBydepth)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
library(agricolae)

TukeyHSD(temp_July_anova)

july_anova_plot <-
  ggplot(lake_july, aes(x = lakename, y = temperature_C, fill = lakename)) +
  geom_boxplot(alpha = 0.8) +
  mytheme +
  labs(
    title = "Mean Temperature Differences Among Lakes (July)",
    x = "Lake Name",
    y = "Temperature (°C)",
    fill = "Lake Name"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 11),
  )

print(july_anova_plot)


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Based on the Tukey’s HSD test results, Peter Lake has statistically similar mean temperatures to Paul Lake and Ward Lake (p > 0.05). All other lakes show significant differences in mean temperature compared to Peter Lake (p < 0.05). Furthermore, no single lake has a mean temperature that is statistically distinct from all the other lakes, indicating that while some lakes differ significantly, there is overlap in temperature groups among several lakes.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can run a Independent two-sample t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
Crampton_ward_july <- lake_july %>%
  filter(lakename == "Crampton Lake"|lakename =="Ward Lake")

CW_ttest <- t.test(data=Crampton_ward_july, temperature_C ~ lakename)
CW_ttest
```

>Answer: As p-value= 0.2649, we failed to reject the null hypothesis that there is no statistically significant difference in mean July temperature between Crampton Lake and Ward Lake.The mean temperatures for the two lakes are statistically equal as the 95% confidence interval includes 0. This result does match my anwer for part 16, which has  0.97 for the p-adjusted value for the two lakes , indicating no significant difference in mean temperatures.
