---
title: "Assignment 3: Data Exploration"
author: "Jingxin Liang"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}

library(tidyverse)
library(lubridate)
library(here)
getwd()
Neonics <- read.csv(here("Data", "Raw", "ECOTOX_Neonicotinoids_Insects_raw.csv"),
                    stringsAsFactors = TRUE)
Litter <- read.csv(here("Data", "Raw", "NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
                   stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We are interested in the ecotoxicology of neonicotinoids on insects because they are widely used agricultural insecticides that can affect not only target pests but also non-target insects such as pollinators. Studying their effects helps us understand the potential ecological cascade — for example, declines in pollinators can disrupt plant reproduction, crop yields, and food webs.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We are interested in studying litter and woody debris because they play a key role in nutrient cycling and soil fertility. As they decompose, they release essential nutrients back into the ecosystem and contribute to soil organic matter. In addition, decomposition rates act as indicators of environmental change, since they are strongly influenced by climate, soil conditions, and microbial activity. Monitoring these processes helps us understand how forests respond to changing environmental conditions.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. NEON collects litter and woody debris in tower plots using 1–4 pairs of elevated and ground traps per plot. Trap locations are chosen randomly when woody cover exceeds 50%, or placed under vegetation when cover is sparse, with additional spacing rules to avoid features like roads, buildings, and streams.
 2. The smallest unit of NEON litterfall sampling is an individual trap, each labeled with a unique trapID. Traps are nested within a hierarchy of identifiers: trapID → subplotID → plotID → siteID → domainID. Spatial information such as latitude, longitude, and elevation of plot centroids is provided along with uncertainty estimates. For higher accuracy, users can either apply the geoNEON R package or use NEON’s API together with a clip cell lookup table to determine exact trap locations.
 3. NEON maintains data quality by applying automated validation checks during data entry and processing prior to release, along with ongoing revisions while data are provisional. Any corrections or updates are recorded in change logs, and known issues are flagged with a quality system—currently limited to a “legacyData” tag for older paper-based records.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
dim(Neonics)
view(data_frame(Neonics))
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}

effect_freq <- sort(table(Neonics$Effect), decreasing = TRUE)


```

> Answer:
The most common effects studied are Population, Mortality, and Behavior. These are of particular interest because mortality measures direct toxic effects, behavior reflects sub-lethal changes that can still threaten survival, and population effects integrate these responses over time to show how neonicotinoid exposure influences long-term persistence of insect populations and the ecosystems they support.



7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
species_freq <- summary(Neonics$Species.Common.Name, maxsum = 7)
species_freq

unique(Neonics$Species.Common.Name)
table(Neonics$Species.Common.Name)

top6_species <- head(sort(table(Neonics$Species.Common.Name), decreasing = TRUE), 6)
head(top6_species)

```

> Answer: The six most commonly studies species in the dataset were Honey Bee, Parasite Wasp Buff, Tailed Bumblebee, Carniolan Honey Bee, Bumble Bee and Italian Honeybee. They are studied because most are pollinators essential for crops and ecosystems, while parasitoid wasps act as natural pest control. Their health directly impacts food security, biodiversity, and agriculture, making them more important than other insects.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}

class(Neonics$'Conc.1..Author.')
```

> Answer: The class of the Conc.1..Author. column is factor. It is not numeric because some entries include text or non-numeric symbols (detection limits, units, or missing value codes), which prevent R from storing the data as purely numeric values.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}
library(ggplot2)

ggplot(data = Neonics, aes(x = Publication.Year)) +
  geom_freqpoly(binwidth = 1, color = "blue") +
  labs(
    title = "Number of Studies by Publication Year",
    x = "Publication Year",
    y = "Number of Studies"
  )


```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}
ggplot(data = Neonics, aes(x = Publication.Year, color= Test.Location)) +
  geom_freqpoly(binwidth = 1) +
  labs(
    title = "Number of Studies by Publication Year by Location",
    x = "Publication Year",
    y = "Number of Studies",
    color = Neonics$Test.Location
  )

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are Lab and Field Natural. Lab studies dominate, especially after the early 2000s, while Field Natural studies are also frequent but show more variability across years, it's most common around 2008. Other locations, such as Field Artificial and Field Undeterminable, are relatively uncommon throughout the time series.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar(fill = "purple") +
  labs(
    title = "Counts of Endpoints in ECOTOX Neonics Dataset",
    x = "Endpoint",
    y = "Count"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

> Answer: The two most common end points are LOEL and NOEL. LOEL is in Terrestrial
database, defined as Lowest -observable -effect-level: lowest dose 
(concentration) producing effects that were significantly different (as reported
by authors) from responses of controls (LOEAL/LOEC).
NOEL: is in Terrestrial database, Defiened as No-observable-effect-level: highest 
dose (concentration)producing effects not significantly different from responses of
controls according to author's reported statistical test
(NOEAL/NOEC)

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
view(data_frame(Litter))
class(Litter$collectDate)
Litter$collectDate <- as.Date(as.character(Litter$collectDate), format = "%Y-%m-%d")
class(Litter$collectDate)

unique(Litter$collectDate)

```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}

unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: unique(Litter$plotID) lists each distinct plotID sampled at Niwot Ridge once, without repetition, while summary(Litter$plotID) shows the frequency of observations for each plotID, providing counts at the same time.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
ggplot(Litter, aes(x = functionalGroup)) +
  geom_bar(fill = 'lightgreen') +
  labs(
    title = "Types of litter in Niwot Ridge sites",
    x = "FunctionalGroup",
    y = "Count"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}

ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Boxplot of Dry Mass by Functional Group",
    x = "Functional Group",
    y = "Dry Mass"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin(fill = "lightblue") +
  labs(
    title = "Violin Plot1 of Dry Mass by Functional Group",
    x = "Functional Group",
    y = "Dry Mass"
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))



```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: A boxplot is more effective than a violin plot in this case because dryMass values are highly skewed, with a few very large outliers. The boxplot clearly shows medians, quartiles, and outliers, while the violin plot becomes distorted and uninformative when extreme values stretch the distribution.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Litters with Needles tend to have the highest biomass at these sites.
