---
title: "Assignment 5: Data Visualization"
author: Jingxin Liang
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
library(tidyverse)
library(here)
library(cowplot)
library(lubridate)

here()
processed_data = here("Data/Processed_KEY")
PeterPaul.chem.nutrients <- read.csv(
  here(processed_data,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

NIWO_Litter_mass <- read.csv(
  here(processed_data,"NEON_NIWO_Litter_mass_trap_Processed.csv"), 
  stringsAsFactors = TRUE)

view(NIWO_Litter_mass)
view(PeterPaul.chem.nutrients)


#2 

NIWO_Litter_mass <- NIWO_Litter_mass %>%
  mutate(collectDate = ymd(collectDate))

PeterPaul.chem.nutrients <- PeterPaul.chem.nutrients %>%
  mutate(collectDate = ymd(sampledate))

class(NIWO_Litter_mass$collectDate)
class(NIWO_Litter_mass$collectDate)
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
# Define your custom theme
theme_mydefault <- theme_minimal(base_size = 13, base_family = "Helvetica") +
  theme(
    # Customize plot background
    plot.background = element_rect(fill = "gray95", color = NA),
    
    # Customize plot title
    plot.title = element_text(
      face = "bold",
      size = 16,
      color = "#C8A2C8",
      hjust = 0.5,        # Center title
      margin = margin(b = 10)
    ),
    
    # Customize axis titles
    axis.title = element_text(face = "bold", color = "#1A5276", size = 12),
    
    # Customize axis ticks and gridlines
    axis.ticks = element_line(color = "gray40"),
    panel.grid.major = element_line(color = "gray80", size = 0.3),
    panel.grid.minor = element_blank(),
    
    # Customize legend
    legend.background = element_rect(fill = "white", color = "gray80"),
    legend.title = element_text(face = "bold", color = "#283747"),
    legend.text = element_text(color = "gray")
  )
theme_set(theme_mydefault)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

#Total P vs PO4 (fixed aes syntax; remove extra comma)
pp_plot_df <- PeterPaul.chem.nutrients %>%
  filter(lakename %in% c("Peter Lake", "Paul Lake")) %>%
  drop_na(tp_ug, po4)

#  checks
nrow(pp_plot_df)
table(pp_plot_df$lakename)

ggplot(pp_plot_df, aes(x = po4, y = tp_ug, color = lakename, shape = lakename)) +
  geom_point(alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15) +   # no fill mapping -> single legend
  coord_cartesian(
    xlim = quantile(pp_plot_df$po4, c(0.01, 0.99), na.rm = TRUE),
    ylim = quantile(pp_plot_df$tp_ug, c(0.01, 0.99), na.rm = TRUE)
  ) +
  labs(
    title = "Total Phosphorus (TP) vs. Phosphate (PO₄)",
    subtitle = "Peter and Paul Lakes (with linear model fits)",
    x = expression(paste("Phosphate (PO"[4], ", ", mu, "g/L)")),
    y = expression(paste("Total Phosphorus (TP, ", mu, "g/L)")),
    color = "Lake",
    shape = "Lake"
  )


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
PeterPaul.chem.nutrients <- PeterPaul.chem.nutrients %>%
  mutate(
    sampledate = as.character(sampledate),                # handle factor -> character
    sampledate = lubridate::parse_date_time(sampledate, orders = c("ymd","mdy","dmy")),
    sampledate = as.Date(sampledate)
  )

pp <- PeterPaul.chem.nutrients %>%
  mutate(
    month = factor(lubridate::month(sampledate, label = TRUE, abbr = TRUE), levels = month.abb),
    lake  = stringr::str_trim(as.character(lakename))
  ) %>%
  filter(lake %in% c("Peter Lake", "Paul Lake"))

nrow(pp)
table(pp$lake)

small_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.background = element_rect(fill = "gray98", color = NA),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    axis.title.x = element_blank()
  )

# 1. Temperature plot
p_temp <- ggplot(pp, aes(x = month, y = temperature_C, color = lake)) +
  geom_boxplot(outlier.size = 1, alpha = 0.6) +
  labs(y = "Temperature (°C)", title = "(a) Temperature") +
  scale_color_brewer(palette = "Set1", name = "Lake") +
  small_theme

# 2. Total Phosphorus plot
p_tp <- ggplot(pp, aes(x = month, y = tp_ug, color = lake)) +
  geom_boxplot(outlier.size = 1, alpha = 0.6) +
  labs(y = expression(paste("TP (", mu, "g/L)")), 
       title = "(b) Total Phosphorus") +
  scale_color_brewer(palette = "Set1", name = "Lake") +
  small_theme

# 3. Total Nitrogen plot 

p_tn <- ggplot(pp, aes(x = month, y = tn_ug, color = lake)) +
  geom_boxplot(outlier.size = 1, alpha = 0.6) +
  labs(y = expression(paste("TN (", mu, "g/L)")), 
       title = "(c) Total Nitrogen") +
  scale_color_brewer(palette = "Set1", name = "Lake") +
  small_theme

legend <- cowplot::get_legend(p_temp + theme(legend.position = "right"))

p_temp_noleg <- p_temp + theme(legend.position = "none")
p_tp_noleg   <- p_tp   + theme(legend.position = "none")
p_tn_noleg   <- p_tn   + theme(legend.position = "none")

plots_col <- cowplot::plot_grid(
  p_temp_noleg, p_tp_noleg, p_tn_noleg,
  ncol = 1, align = "v", axis = "lr", rel_heights = c(1, 1, 1)
)

final_plot <- cowplot::plot_grid(plots_col, legend, ncol = 2, rel_widths = c(1, 0.25))

print(final_plot)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: I observed that temperature strongly reflects seasonal changes, while nutrient concentrations (TP and TN) show greater variability between lakes. Peter Lake tends to have higher nutrient levels than Paul Lake, particularly for nitrogen, indicating differing nutrient dynamics or watershed characteristics between the two systems.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
# ensure collectDate is Date (safe conversion)
NIWO_Litter_mass <- NIWO_Litter_mass %>%
  mutate(
    collectDate = as.character(collectDate),
    collectDate = lubridate::parse_date_time(collectDate, orders = c("ymd","mdy","dmy")),
    collectDate = as.Date(collectDate)
  )

# inspect functional groups (optional)
unique(NIWO_Litter_mass$functionalGroup)
needles <- NIWO_Litter_mass %>%
  filter(!is.na(collectDate) & !is.na(dryMass)) %>%
  filter(str_to_lower(functionalGroup) == "needles")

# quick check: NLCD classes present
unique(needles$nlcdClass)
p_needles_color <- ggplot(needles, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(aes(group = nlcdClass), method = "loess", se = FALSE, linetype = "solid", size = 0.6) +
  scale_color_brewer(palette = "Set2", name = "NLCD class") +
  labs(
    title = "Needle litter dry mass through time",
    x = "Date (month)",
    y = "Dry mass (g)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )
 
print(p_needles_color)

#7

p_needles_facets <- ggplot(needles, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point(alpha = 0.6, size = 1.6) +
  geom_smooth(method = "loess", se = FALSE, size = 0.7) +
  facet_wrap(~ nlcdClass, ncol = 3, scales = "free_y") +
  scale_color_brewer(palette = "Set2", name = "NLCD class") +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "6 months",
               expand = expansion(mult = c(0.01, 0.01))) +
  labs(
    title = "Needle litter dry mass by NLCD class (faceted, colored by class)",
    x = "Date",
    y = "Dry mass (g)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"   # hide redundant legend if desired
  )

print(p_needles_facets)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Between the two plots, the faceted plot (Plot 7) is more effective in presenting the data. By separating each NLCD class into its own panel, it avoids the overlap of points and trend lines that makes the combined plot (Plot 6) harder to interpret. This layout allows for clearer visualization of the temporal patterns within each land cover type and makes it easier to compare differences in dry mass trends across classes. The consistent axes across panels also support direct visual comparison. In contrast, the combined plot is more cluttered and better suited only for showing overall patterns rather than detailed class-specific trends.