---
title: "Assignment 2: Coding Basics"
author: Jingxin Liang
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 55, increasing by fives. Assign this sequence a name.

Sequen_1to55 <- seq(from=1, to=55,  by=5)

2.  Compute the mean and median of this sequence.

Ave_seq <- mean(Sequen_1to55)
Ave_seq

Median <- median(Sequen_1to55)
Median

3.  Ask R to determine whether the mean is greater than the median.

Median < Ave_seq

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Create a sequence of numbers from 1 to 55, increasing by 5
Sequen_1to55 <- seq(from=1, to=55,  by=5)

#2. Calculate the mean and median of the sequence
Ave_seq <- mean(Sequen_1to55)
Ave_seq

Median <- median(Sequen_1to55)
Median

#3.Compare whether the median is less than the mean (returns TRUE or FALSE)
Median < Ave_seq

```

## Basics, Part 2

5.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

#(a) student names, character vector
students <- c("Michelle", "Isabella", "Bella", "Molly")

# (b) test scores, numeric vector
scores <- c(85, 92, 88, 76)

# (c) scholarship status, logical vector
scholarship<- c(FALSE, TRUE, TRUE, FALSE)
scholarship

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

student_data <- data.frame(students, scores, scholarship)
student_data

8.  Label the columns of your data frame with informative titles.

student_data <- data.frame(Name=students, Scores=scores, Scholarship=scholarship)
student_data

```{r basics 2}
#(a) student names, character vector
students <- c("Michelle", "Isabella", "Bella", "Molly")

# (b) test scores, numeric vector
scores <- c(85, 92, 88, 76)

# (c) scholarship status, logical vector
scholarship<- c(FALSE, TRUE, TRUE, FALSE)
scholarship

student_data <- data.frame(students, scores, scholarship)
student_data

student_data <- data.frame(Name=students, Scores=scores, Scholarship=scholarship)
student_data

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Matrix can only have one data type while data frame can have multiple different data types.

10. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, print the word "Pass"; otherwise print the word "Fail". 

greater_than5_1 <- function(number) {
  if ( number > 50) {
    print("Pass")}
 else {
    print("Fail")
  }
}
greater_than5_1(52.5)


11. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

greater_than5_2 <- function(number) {
  ifelse(number > 50, "Pass", "Fail")}
greater_than5_2(52.5)  

12. Run both functions using the value 52.5 as the input

13. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

#greater_than5_1 <- function(number) {
#  if ( number > 50) {
#    print("Pass")}
# else {
#    print("Fail")}}
  

#greater_than5_1(scores)

greater_than5_2 <- function(number) {
  ifelse(number > 50, "Pass", "Fail")
}
greater_than5_2(scores)


```{r basics 3}
#10. Create a function using if...else

greater_than5_1 <- function(number) {
  if ( number > 50) {
    print("Pass")}
 else {
    print("Fail")
  }
}
greater_than5_1(52.5)


#11. Create a function using ifelse()

greater_than5_2 <- function(number) {
  ifelse(number > 50, "Pass", "Fail")}

greater_than5_2(52.5)

#12a. Run the first function with the value 52.5
greater_than5_1(52.5) 
#12b. Run the second function with the value 52.5
greater_than5_2(52.5) 
#13a. Run the first function with the vector of test scores
#greater_than5_1 <- function(number) {
#  if ( number > 50) {
#    print("Pass")}
# else {
#    print("Fail")}}
  

#greater_than5_1(scores)


#13b. Run the second function with the vector of test scores
greater_than5_2 <- function(number) {
  ifelse(number > 50, "Pass", "Fail")
}
greater_than5_2(scores)
```

14. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization")
 
> Answer: The ifelse worked, since  "if..else" can only have one condition.


**NOTE** Before knitting, you'll need to comment out the call to the function in Q13 that does not work. (A document can't knit if the code it contains causes an error!)